Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
case study and methods structrure

---
title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR"
author:
  - name: Rosa Félix\corref{cor1}
    affiliation: CERIS
    correspondingauthor: true
    # footnote: 1
    email: rosamfelix@tecnico.ulisboa.pt
  - name: Filipe Moura
    affiliation: CERIS
    email: fmoura@tecnico.ulisboa.pt
  - name: Robin Lovelace
    affiliation: ITS
    email: r.lovelace@uleeds.uk
address:
  - code: CERIS
    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
  - code: ITS
    address: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
abstract: |
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. The impacts of shifting car trips to public transport were also estimated and monetized.
  The results show that 10% of all trips could be made by cycling in combination with public transport. Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year. The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
  \
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
journal: "Computers, Environment and Urban Studies"
date: "`r Sys.Date()`"
classoption: review, doubleblind, 3p, authoryear #change later to preprint instead of review
bibliography: bibliography.bib
biblio-style: elsarticle-harv
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/computers-environment-and-urban-systems
# csl: computers-environment-and-urban-systems.csl
preamble: |
  \biboptions{authoryear}
  \usepackage{url}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{makecell}
  \usepackage{multirow}
output: 
  rticles::elsevier_article:
    keep_tex: yes
    # citation_package: "default" #ou nenhum
    latex_engine: xelatex
    extra_dependencies: "subfig"
#   bookdown::word_document2:
#     fig_caption: true
# always_allow_html: true
urlcolor: blue
citecolor: black
linkcolor: blue
# natbiboptions: longnamesfirst,angle,semicolon
---
\hypersetup{citecolor=black}

@@

The potential of cycling as a complementary mode of PT is substantial worldwide, especially in cities with established public transport networks or substantial ambitions to develop them.
In the Lisbon metropolitan area (LMA) the largest metropolitan area in Portugal, the modal share of cycling is low, but the potential for cycling as a complementary mode of PT is high.
According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
Car modal share was 58.4%, while PT accounted for 15.5% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in the same municipality --- amounts to 3.5 million trips.
This exceeds the number of inter-municipal trips (1.8 million trips), involving travel between different municipalities.
Cars and public transport are the most used modes for intercity trips, with cars being the predominant choice for all journeys.
`r round(100*2948314/5569731)`% trips are up to 5 km distance, and `r round(100*3958667/5569731)`% up to 10 km.
Nevertheless, 29% of trips are longer than 10 km, which requires the use of motorized modes, or active modes in combination with public transport.

@@

It is essential to note that our approach is not predictive but scenario-based, providing a range of potential outcomes regarding the target (% of cycling tips) and the routing profile (safe or direct routes), rather than a single forecast.
These scenarios are rooted in the explicit national cycling targets, assuming that the potential for cycling is proportional to the reduction of car trips.
Thus said, we do not predict the generation of new trips, but the substitution of car trips by cycling in combination with PT, considering the potential of the first and last mile of these journeys to be made by bicycle.
Additionally, we do not model changes in the baseline PT trips, under the implicit assumption that the initial and final stages of these trips remain unaltered.
A strength of this _backcasting_ approach is that we can simulate the aggregate travel patterns that would meet particular policy objectives, and explore what the implications could be for the public transport system and provision of safe cycling infrastructure to public transport nodes at regional, local authority and, importantly from a local transport planning practitioner perspective, corridor level.

@@

# od_all = readRDS("paper/paperTRA/load/od_all.Rds")
# od_jittered_filter = readRDS("paper/paperTRA/load/od_jittered_filter.Rds")
# 
# odcombined = rbind(
#   od_all |>
#     transmute(Total, Disaggregation = " Original"),
#   od_jittered_filter |>
#     transmute(Total, Disaggregation = "Jittered")
# )
# 
# cairo_pdf(filename = "img/jitter_cairo.pdf", onefile= T, width=10, height=4) 
# 
# odcombined |>
#   sample_n(10000, weight = odcombined$Total) |>
#   mutate(Total = Total / 1000) |>
#   ggplot() +
#   geom_sf(aes(alpha = Total)) + # blured
#   facet_wrap(~Disaggregation) +
#   theme_void() +
#   theme(legend.position="none",
#         strip.text.x = element_blank()) +
#   geom_sf(data = MUNICIPIOSgeo, fill = NA, color = "gray60")
# 
# dev.off()
# 
# png::writePNG(pdftools::pdf_render_page("img/jitter_cairo.pdf", dpi = 300), "img/jitter_cairo.png")

@@

@@

This paper focus primarily on the potential for replacing trips currently made by car with combined cycling and public transport trips, and the associated social and environmental benefits.
While the results are specific to the Lisbon Metropolitan Area, the approach is designed to be generalizable to other metropolitan areas.

@@

Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.
For instance, we considered only the inter-municipal bus services that accommodate bicycles on board.
Future research could explore the potential of equipping other regional bus services with this feature and analyse their attractiveness, in conjunction with cycling, to address current car trip demands.

A strength of the approach is that there is no requirement for sensitive individual trip level data to reproduce the approach in other cities.
However, we acknowledge that this approach has limitations.
The approach cannot show exactly which trips are being replaced, in terms of the age of people making the trips, the types of car they drive, and other individual or vehicle level variables.
We assumed uniformity in the type of vehicle for all replaced car journeys, albeit with variations in the fuels utilized.
Additionally, we assumed consistency in the health impacts across all individuals for the replaced car trips.
This approach is a simplification of reality to enable the analysis of a large region on available computing resources.
Future work could seek to increase the resolution of the results, and move towards a more individual-level analysis, for example by using agent and activity-based models using tools such as MATSim [@MATSim].
This direction of travel would require substantial resources outside the scope of this study; another way to improve the spatial resolution of results would be to use higher resolution 'subpoints' as the basis of the desire line 'jittering' process.
We would also like to explore the potential for probabilistic routing to improve the spatial resolution of the results.

A further refinement could be to differentiate between 'simpler' no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods, by relaxing the assumption of no-transfers between PT modes in the routing engine.
This would led to a larger number of total daily trips that can be made by cycling + PT, in particular in locations where the PT network is less complex and requires more transfers.
We could also set a different maximum lenght of cycling stage, extending the bicycle catchment area.
However, despite the added value of a sensibility analyses, they would also increase the overall complexity, and would have an impact on the explainability of the results to be used by policy makers.

>>>>

---
title: "Reproducible methods for modeling combined public transport and cycling trips and associated benefits: \ evidence from the biclaR tool"
author:
  - name: Rosa Félix\corref{cor1}
    affiliation: CERIS
    correspondingauthor: true
    # footnote: 1
    email: rosamfelix@tecnico.ulisboa.pt
  - name: Filipe Moura
    affiliation: CERIS
    email: fmoura@tecnico.ulisboa.pt
  - name: Robin Lovelace
    affiliation: ITS
    email: r.lovelace@uleeds.uk
address:
  - code: CERIS
    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
  - code: ITS
    address: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
abstract: |
  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This paper estimates the potential for cycling combined with public transit as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_). The social and environmental impacts were assessed using Health World Organization tools. The impacts of shifting car trips to public transport were also estimated and monetized.
  The results show that 10% of all trips could be made by cycling in combination with public transport. Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year. The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
  \
keywords: 
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods
journal: "Computers, Environment and Urban Studies"
date: "`r Sys.Date()`"
classoption: review, doubleblind, 3p, authoryear #change later to preprint instead of review
bibliography: bibliography.bib
biblio-style: elsarticle-harv
linenumbers: true
numbersections: true
# Use a CSL with `citation_package = "default"`
csl: https://www.zotero.org/styles/computers-environment-and-urban-systems
# csl: computers-environment-and-urban-systems.csl
preamble: |
  \biboptions{authoryear}
  \usepackage{url}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{longtable}
  \usepackage{makecell}
  \usepackage{multirow}
output: 
  rticles::elsevier_article:
    keep_tex: yes
    # citation_package: "default" #ou nenhum
    latex_engine: xelatex
    extra_dependencies: "subfig"
#   bookdown::word_document2:
#     fig_caption: true
# always_allow_html: true
urlcolor: blue
citecolor: black
linkcolor: blue
# natbiboptions: longnamesfirst,angle,semicolon
---
\hypersetup{citecolor=black}

@@

The potential of cycling as a complementary mode of PT is substantial worldwide, especially in cities with established public transport networks or substantial ambitions to develop them.
In the Lisbon metropolitan area (LMA) the largest metropolitan area in Portugal, the modal share of cycling is low, but the potential for cycling as a complementary mode of PT is high.

According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
Car modal share was 58.4%, while PT accounted for 15.5% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in the same municipality --- amounts to 3.5 million trips.
This exceeds the number of inter-municipal trips (1.8 million trips), involving travel between different municipalities.
Cars and public transport are the most used modes for intercity trips, with cars being the predominant choice for all journeys.
`r round(100*2948314/5569731)`% trips are up to 5 km distance, and `r round(100*3958667/5569731)`% up to 10 km.
Nevertheless, 29% of trips are longer than 10 km, which requires the use of motorized modes, or active modes in combination with public transport.

@@

@@

# od_all = readRDS("paper/paperTRA/load/od_all.Rds")
# od_jittered_filter = readRDS("paper/paperTRA/load/od_jittered_filter.Rds")
# MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
#
# odcombined = rbind(
#   od_all |>
#     transmute(Total, Disaggregation = " Original"),
#   od_jittered_filter |>
#     transmute(Total, Disaggregation = "Jittered")
# )
# 
# cairo_pdf(filename = "img/jitter_cairo.pdf", onefile= T, width=10, height=4) 
# 
# odcombined |>
#   sample_n(10000, weight = odcombined$Total) |>
#   mutate(Total = Total / 1000) |>
#   ggplot() +
#   geom_sf(aes(alpha = Total)) + # blured
#   facet_wrap(~Disaggregation) +
#   theme_void() +
#   theme(legend.position="none",
#         strip.text.x = element_blank()) +
#   geom_sf(data = MUNICIPIOSgeo, fill = NA, color = "gray60")
# 
# dev.off()
# 
# png::writePNG(pdftools::pdf_render_page("img/jitter_cairo.pdf", dpi = 300), "img/jitter_cairo.png")

@@

## Modelling car shift to bicycle-PT - here or before??

It is essential to note that our approach is not predictive but scenario-based, providing a range of potential outcomes regarding the target (% of cycling tips) and the routing profile (safe or direct routes), rather than a single forecast.
These scenarios are rooted in the explicit national cycling targets, assuming that the potential for cycling is proportional to the reduction of car trips.
Thus said, we do not predict the generation of new trips, but the substitution of car trips by cycling in combination with PT, considering the potential of the first and last mile of these journeys to be made by bicycle.
Additionally, we do not model changes in the baseline PT trips, under the implicit assumption that the initial and final stages of these trips remain unaltered.

A strength of this _backcasting_ approach is that we can simulate the aggregate travel patterns that would meet particular policy objectives, and explore what the implications could be for the public transport system and provision of safe cycling infrastructure to public transport nodes at regional, local authority and, importantly from a local transport planning practitioner perspective, corridor level.

@@

This paper focus primarily on the potential for replacing trips currently made by car with combined cycling and public transport trips, and the associated social and environmental benefits, showcasing the potential impact of facilitating the first-and-last mile to PT interfaces.
While the results are specific to the Lisbon Metropolitan Area, the approach is designed to be generalizable to other metropolitan areas.

@@

Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.
For instance, we considered only the inter-municipal bus services that accommodate bicycles on board.
Future research could explore the potential of equipping other regional bus services with this feature and analyse their attractiveness, combined with cycling, to address current car trip demands.

A strength of the approach is that there is no requirement for sensitive individual trip level data to reproduce the approach in other cities.
However, we acknowledge that this approach has limitations.
The approach cannot show exactly which trips are being replaced, in terms of the age of people making the trips, the types of car they drive, and other individual or vehicle level variables.
We assumed uniformity in the type of vehicle for all replaced car journeys, albeit with variations in the fuels utilised.
Additionally, we assumed consistency in the health impacts across all individuals for the replaced car trips.
This approach is a simplification of reality to enable the analysis of a large region on available computing resources.
Future work could seek to increase the resolution of the results, and move towards a more individual-level analysis, for example by using agent and activity-based models using tools such as MATSim [@MATSim].
This option would require substantial resources outside the scope of this study; another way to improve the spatial resolution of results would be to use higher resolution 'subpoints' as the basis of the desire line 'jittering' process.
We would also like to explore the potential for probabilistic routing to improve the spatial resolution of the results.

A further refinement could be to differentiate between 'simpler' no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods, by relaxing the assumption of no-transfers between PT modes in the routing engine.
This would lead to a larger number of total daily trips that can be made by cycling + PT, in particular in locations where the PT network is less complex and requires more transfers.
We could also set a different maximum length of cycling stage, extending the bicycle catchment area.
However, despite the added value of a sensitivity analyses, they would also increase the overall complexity, and would have an impact on the explainability of the results to be used by policy makers.

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
case study and methods structrure

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

  \title{Modeling the impacts of replacing car trips with combined
public transport and cycling: Reproducible methods, results and
actionable evidence from biclaR}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  A high proportion of car trips can be replaced by a combination of
  public transit and cycling for the first-and-last mile. This paper
  estimates the potential for cycling combined with public transit as a
  substitute for car trips in the Lisbon metropolitan area and assesses
  its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of
  a metropolitan cycling network was developed (\emph{biclaR}). The
  social and environmental impacts were assessed using the \emph{HEAT
  for Cycling} and the \emph{HEAT as a Service} tools. The impacts of
  shifting car trips to public transport were also estimated and
  monetized. The results show that 10\% of all trips could be made by
  cycling in combination with public transport. Shifting to cycling for
  the shorter first and last mile stages can reduce annual
  CO\textsubscript{2}eq emissions from 3,000 to 7,500 tons/day, while
  for the public transport leg, the transfer from car avoids of up to
  20,500 tons of CO\textsubscript{2}eq emissions per year. The estimated
  socio-environmental benefits are of €125 million to €325 million over
  10 years. This evidence can support policymakers to prioritize
  interventions that reduce the reliance on private motor vehicles.\\
  \end{abstract}
    \begin{keyword}
    Active transport \sep Intermodality \sep First and last
mile \sep Health economic assessment \sep Environmental impacts \sep 
    Open data and methods
  \end{keyword}

@@

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high. According to the latest mobility
survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
million daily trips, with only 0.5\% by bicycle. Car modal share was
58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
km. Nevertheless, 29\% of trips are longer than 10 km, which requires
the use of motorized modes, or active modes in combination with public
transport.

@@

It is essential to note that our approach is not predictive but
scenario-based, providing a range of potential outcomes regarding the
target (\% of cycling tips) and the routing profile (safe or direct
routes), rather than a single forecast. These scenarios are rooted in
the explicit national cycling targets, assuming that the potential for
cycling is proportional to the reduction of car trips. Thus said, we do
not predict the generation of new trips, but the substitution of car
trips by cycling in combination with PT, considering the potential of
the first and last mile of these journeys to be made by bicycle.
Additionally, we do not model changes in the baseline PT trips, under
the implicit assumption that the initial and final stages of these trips
remain unaltered. A strength of this \emph{backcasting} approach is that
we can simulate the aggregate travel patterns that would meet particular
policy objectives, and explore what the implications could be for the
public transport system and provision of safe cycling infrastructure to
public transport nodes at regional, local authority and, importantly
from a local transport planning practitioner perspective, corridor
level.

@@

@@

Nevertheless, the extent of the potential shift from car to PT is highly
dependent on the distances typically travelled in such contexts, and the
availability of PT services - which can vary significantly between
regions, and for which we are not assuming any change in this study. For
instance, we considered only the inter-municipal bus services that
accommodate bicycles on board. Future research could explore the
potential of equipping other regional bus services with this feature and
analyse their attractiveness, in conjunction with cycling, to address
current car trip demands.

A strength of the approach is that there is no requirement for sensitive
individual trip level data to reproduce the approach in other cities.
However, we acknowledge that this approach has limitations. The approach
cannot show exactly which trips are being replaced, in terms of the age
of people making the trips, the types of car they drive, and other
individual or vehicle level variables. We assumed uniformity in the type
of vehicle for all replaced car journeys, albeit with variations in the
fuels utilized. Additionally, we assumed consistency in the health
impacts across all individuals for the replaced car trips. This approach
is a simplification of reality to enable the analysis of a large region
on available computing resources. Future work could seek to increase the
resolution of the results, and move towards a more individual-level
analysis, for example by using agent and activity-based models using
tools such as MATSim \citep{MATSim}. This direction of travel would
require substantial resources outside the scope of this study; another
way to improve the spatial resolution of results would be to use higher
resolution `subpoints' as the basis of the desire line `jittering'
process. We would also like to explore the potential for probabilistic
routing to improve the spatial resolution of the results.

A further refinement could be to differentiate between `simpler'
no-transfer PT trips (which require less person effort) and more arduous
multi-transfer PT trips, using the same methods, by relaxing the
assumption of no-transfers between PT modes in the routing engine. This
would led to a larger number of total daily trips that can be made by
cycling + PT, in particular in locations where the PT network is less
complex and requires more transfers. We could also set a different
maximum lenght of cycling stage, extending the bicycle catchment area.
However, despite the added value of a sensibility analyses, they would
also increase the overall complexity, and would have an impact on the
explainability of the results to be used by policy makers.

>>>>

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

@@

  \title{Reproducible methods for modeling combined public transport and
cycling trips and associated benefits: evidence from the biclaR tool}
    \author[CERIS]{Rosa Félix\corref{cor1}%
  \corref{cor1}%
  }
   \ead{rosamfelix@tecnico.ulisboa.pt} 
    \author[CERIS]{Filipe Moura%
  %
  }
   \ead{fmoura@tecnico.ulisboa.pt} 
    \author[ITS]{Robin Lovelace%
  %
  }
   \ead{r.lovelace@uleeds.uk} 
      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
    \affiliation[ITS]{Institute for Transport Studies, University of
Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
    \cortext[cor1]{Corresponding author}

  \begin{abstract}
  A high proportion of car trips can be replaced by a combination of
  public transit and cycling for the first-and-last mile. This paper
  estimates the potential for cycling combined with public transit as a
  substitute for car trips in the Lisbon metropolitan area and assesses
  its socio-environmental impacts using open data and open source tools.
  A decision support tool that facilitates the design and development of
  a metropolitan cycling network was developed (\emph{biclaR}). The
  social and environmental impacts were assessed using Health World
  Organization tools. The impacts of shifting car trips to public
  transport were also estimated and monetized. The results show that
  10\% of all trips could be made by cycling in combination with public
  transport. Shifting to cycling for the shorter first and last mile
  stages can reduce annual CO\textsubscript{2}eq emissions from 3,000 to
  7,500 tons/day, while for the public transport leg, the transfer from
  car avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
  year. The estimated socio-environmental benefits are of €125 million
  to €325 million over 10 years. This evidence can support policymakers
  to prioritize interventions that reduce the reliance on private motor
  vehicles.\\
  \end{abstract}
    \begin{keyword}
    Active transport \sep Intermodality \sep First and last
mile \sep Health economic assessment \sep Environmental impacts \sep 
    Open data and methods
  \end{keyword}

@@

The potential of cycling as a complementary mode of PT is substantial
worldwide, especially in cities with established public transport
networks or substantial ambitions to develop them. In the Lisbon
metropolitan area (LMA) the largest metropolitan area in Portugal, the
modal share of cycling is low, but the potential for cycling as a
complementary mode of PT is high.

According to the latest mobility survey conducted in 2018 \citep{IMOB},
the LMA registered a total of 5.3 million daily trips, with only 0.5\%
by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%
(see Figure \ref{fig:mododist}). The number of intra-municipal trips ---
with origin and destination in the same municipality --- amounts to 3.5
million trips. This exceeds the number of inter-municipal trips (1.8
million trips), involving travel between different municipalities. Cars
and public transport are the most used modes for intercity trips, with
cars being the predominant choice for all journeys. 53\% trips are up to
5 km distance, and 71\% up to 10 km. Nevertheless, 29\% of trips are
longer than 10 km, which requires the use of motorized modes, or active
modes in combination with public transport.

@@

@@

\subsection{Modelling car shift to bicycle-PT - here or
before??}\label{modelling-car-shift-to-bicycle-pt---here-or-before}

It is essential to note that our approach is not predictive but
scenario-based, providing a range of potential outcomes regarding the
target (\% of cycling tips) and the routing profile (safe or direct
routes), rather than a single forecast. These scenarios are rooted in
the explicit national cycling targets, assuming that the potential for
cycling is proportional to the reduction of car trips. Thus said, we do
not predict the generation of new trips, but the substitution of car
trips by cycling in combination with PT, considering the potential of
the first and last mile of these journeys to be made by bicycle.
Additionally, we do not model changes in the baseline PT trips, under
the implicit assumption that the initial and final stages of these trips
remain unaltered.

A strength of this \emph{backcasting} approach is that we can simulate
the aggregate travel patterns that would meet particular policy
objectives, and explore what the implications could be for the public
transport system and provision of safe cycling infrastructure to public
transport nodes at regional, local authority and, importantly from a
local transport planning practitioner perspective, corridor level.

@@

Nevertheless, the extent of the potential shift from car to PT is highly
dependent on the distances typically travelled in such contexts, and the
availability of PT services - which can vary significantly between
regions, and for which we are not assuming any change in this study. For
instance, we considered only the inter-municipal bus services that
accommodate bicycles on board. Future research could explore the
potential of equipping other regional bus services with this feature and
analyse their attractiveness, combined with cycling, to address current
car trip demands.

A strength of the approach is that there is no requirement for sensitive
individual trip level data to reproduce the approach in other cities.
However, we acknowledge that this approach has limitations. The approach
cannot show exactly which trips are being replaced, in terms of the age
of people making the trips, the types of car they drive, and other
individual or vehicle level variables. We assumed uniformity in the type
of vehicle for all replaced car journeys, albeit with variations in the
fuels utilised. Additionally, we assumed consistency in the health
impacts across all individuals for the replaced car trips. This approach
is a simplification of reality to enable the analysis of a large region
on available computing resources. Future work could seek to increase the
resolution of the results, and move towards a more individual-level
analysis, for example by using agent and activity-based models using
tools such as MATSim \citep{MATSim}. This option would require
substantial resources outside the scope of this study; another way to
improve the spatial resolution of results would be to use higher
resolution `subpoints' as the basis of the desire line `jittering'
process. We would also like to explore the potential for probabilistic
routing to improve the spatial resolution of the results.

A further refinement could be to differentiate between `simpler'
no-transfer PT trips (which require less person effort) and more arduous
multi-transfer PT trips, using the same methods, by relaxing the
assumption of no-transfers between PT modes in the routing engine. This
would lead to a larger number of total daily trips that can be made by
cycling + PT, in particular in locations where the PT network is less
complex and requires more transfers. We could also set a different
maximum length of cycling stage, extending the bicycle catchment area.
However, despite the added value of a sensitivity analyses, they would
also increase the overall complexity, and would have an impact on the
explainability of the results to be used by policy makers.
